YALE DEPARTMENT OF ECONOMICS \\ \vspace*{60pt} The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-30 draft.Rnw
Wrote conclusion.

\subsection{ADDED}
The finding that the features that are selected to predict superior returns
through my purely mechanical procedure end up being very similar to the features
subsequently used by Prosper to set credit grades supports the view
that investors failed to attend to these variables, as Freedman and Jin suggest,
and that Prosper as a market maker fixed their errors. Prosper's subsequent
reliance on the information I use to beat the market, supports the argument that
the information was, in fact valuable, while the facts of Section 4, namely
Figure \ref{ratevscore}, aim to show
that the information was largely ignored. Even more importantly, this analysis is NOT
vulnerable to the critique that my tests of market efficiency are undermined by
an unlucky time period. In particular, Prosper used these same variables to
guide investors, and improve market efficiency, at a very different time.

@@

\subsection{Results \& Discussion}
<<adjfig, cache=FALSE, fig.cap='This chart displays the results of our rate-adjustment experiment. The blue line is the result of our rate-adjustment experiment, and displays the hypothetical ROI if late auction regime borrowers had paid early auction interest rates. The grey cloud represents a 2\% adjustment for cost-induced default.'>>=
ggplot(decomp, aes(x=CreditScoreEst, y=meanRoi, color=regime)) +
  #geom_line(aes(x=CreditScoreEst, y=meanRoi, color=regime, size=num)) +
  geom_point(aes(size=num))+
  geom_smooth(aes(ymax=meanRoi, ymin=(meanRoi - err)),stat='identity') +
  theme_bw() +
  labs(x='Estimated Credit Score', color='Regime', y='Mean ROI',size='Number of Loans', title='Rate Adjustment Results')+
  xlim(640,900)
@
Figure \ref{fig:adjfig} documents the relationship between credit score bucket and returns for early auction regime, late auction regime, and late auction regime borrowers with adjusted rates, where the rate adjustment and subsequent ROI adjustment was computed using the \emph{upper bound} methodology described above. At low credit scores, where the difference in interest rates between the two regimes is largest, the gap in ROI is also the largest. With adjusted rates, riskier assets have higher average return than less risky assets, and average returns are positive, suggesting that part of the reason the actual data is inconsistent with the EMH predictions may be the low interest rates and exposure to subprime borrowers.

@@

\section{Conclusion (TODO)}
\begin{enumerate}
\item Restate results.
\item Market maker intervention may be useful when info gathering costs are high.
\item Platform invest in educating investors? (VK idea)
\item Future research: same experiment with known ROI. Better time controls? Look at consistent errors in mistakes borrowers made.
\end{enumerate}
\newpage
\section{Appendix}
\subsection{Prosper's Charged-Off vs. Default Terminology}
After the shutdown, Prosper began using the term ``Charged-Off" to mean that a loan is delinquent past 120 days. We observe more defaults than charge-offs before the shutdown, but more charge-offs than defaults after, likely because of the shift in terminology. According to Prosper's website:
\begin{quote}
In general, a debt or account is considered “charged off” when it is unlikely that further payments will be received. Debts are usually charged off after they remain unpaid for a period of time (e.g., 90 to 180 days). Prosper uses the 120 days as the charge off threshold because loans that become over 120 days past due are eligible for sale to a debt buyer, and we have found that there is a steep drop-off in likelihood of further payments after 120 days of delinquency.\footcite{chargeoff}
\end{quote}
\subsection{Out of Sample Performance by Credit Grade}
<<retTab2,cache=FALSE>>=
vmeans = res %>% group_by(regime,model,grade) %>%  
  summarise(mn = sum((1+meanUL) *length) / sum(length) -1, n=sum(length))

>>>>

\subsection{ADDED}
The finding that the features that are selected to predict superior returns
through my purely mechanical procedure end up being very similar to the features
subsequently used by Prosper to set credit grades supports the view
that investors failed to attend to these variables, as Freedman and Jin suggest,
and that Prosper as a market maker fixed their errors. Prosper's subsequent
reliance on the information I use to beat the market, supports the argument that
the information was, in fact valuable, while the facts of Section 4, namely
Figure \ref{fig:ratevscore}, aim to show
that the information was largely ignored. Even more importantly, this analysis is NOT
vulnerable to the critique that my tests of market efficiency are undermined by
an unlucky time period. In particular, Prosper used these same variables to
guide investors, and improve market efficiency, at a very different time.

@@

\subsection{Results \& Discussion}
<<adjfig, cache=FALSE, fig.cap='This chart displays the results of our rate-adjustment experiment. The blue line is the result of our rate-adjustment experiment, and displays the hypothetical ROI if late auction regime borrowers had paid early auction interest rates. The grey cloud represents a 2 percent adjustment for cost-induced default.'>>=
ggplot(decomp, aes(x=CreditScoreEst, y=meanRoi, color=regime)) +
  #geom_line(aes(x=CreditScoreEst, y=meanRoi, color=regime, size=num)) +
  geom_point(aes(size=num))+
  geom_smooth(aes(ymax=meanRoi, ymin=(meanRoi - err)),stat='identity') +
  theme_bw() +
  labs(x='Estimated Credit Score', color='Regime', y='Mean ROI',size='Number of Loans', title='Rate Adjustment Results')+
  xlim(640,900)
@
Figure \ref{fig:adjfig} documents the relationship between credit score bucket and returns for early auction regime, late auction regime, and late auction regime borrowers with adjusted rates, where the rate adjustment and subsequent ROI adjustment was computed using the \emph{upper bound} methodology described above. At low credit scores, where the difference in interest rates between the two regimes is largest, the gap in ROI is also the largest. With adjusted rates, riskier assets have higher average return than less risky assets, and average returns are positive, suggesting that part of the reason the actual data is inconsistent with the EMH predictions may be the low interest rates and exposure to subprime borrowers.

@@

\section{Conclusion}
This paper examined interest rate setting and lender returns on Prosper.com. It
took advantage of three policy regimes over the course of 2006-2011. In the
first regime, investors set interest rates through a decentralized reverse
auction; in the second regime they continued to use the auction, but with
considerably more guidance from Prosper; and in the third regime Prosper set the
rates directly. I find that compared with the later two periods, interest rates
during the early auction regime were much lower and less sensitive to risk, and
that investor returns were much lower than in the later two regimes.

I further examined these findings and argued that the evidence from the early
auction regime is inconsistent with the predictions of the EMH. In particular, I
reported three findings: average returns in the early auction period are
negative, riskier loans earn lower average returns than more risky ones, and
--most importantly -- one can use loan selection strategies without look-ahead
bias to select well-performing loans in the early auction period. The third
finding is inconsistent with the most fundamental prediction of the EMH, that an
investor cannot earn superior risk-adjusted returns based on publicly available
information.

The first two findings are vulnerable to the critique that my results depend on
the particularly bad time period (2006-2008) during the early auction years. I
show that, even correcting for the time period, returns would have been poor
because interest rates were set too low and insufficiently sensitive to risk.
My third finding is more difficult to criticize based on the time period, since
it shows that investors could have done better in the same period, the second
half of the early auction regime, based on publicly available information.

I offer two critical pieces of corroborating evidence. First, when Prosper began
setting interest rates, it appears to haved use publicly available
variables to set interest rates that are very similar to those selected by my
Feature Selection algorithm totally independently, and then used to earn
superior returns by taking advantage of the fact that investors neglected them.
Second, I present some evidence that, in the early auction period, Prosper loans
performed considerably worse than Capital One loans with similar risk
characteristics. In short, all the evidence suggests that the decentralized
unguided mechanism of the early auction period did not lead to efficient
interest rates for Prosper loans.

The last finding has implications for financial economics and market design. It
suggests that contrary to Hayek's vision that the wisdom of the crowds leads to
efficient price setting, the crowds did not utilize all the public information
available to them, and made significant mistakes. Prosper learned this quite
quickly, and intervened in completely decentralized price formation. Guidance,
and eventually price intervention from the market maker, likely improved returns
substantially. The evidence cautions against the assumption of universal
investor rationality. More important, it suggests that design of markets that
protects investors from their errors and limited attention, either by fairly
aggressive guidance or even price setting, can improve market performance
dramatically.

\begin{enumerate}
\item Restate results.
\item Market maker intervention may be useful when info gathering costs are high.
\item Platform invest in educating investors? (VK idea)
\item Future research: same experiment with known ROI. Better time controls? Look at consistent errors in mistakes borrowers made.
\end{enumerate}
\newpage
\section{Appendix}
\subsection{Prosper's Charged-Off vs. Default Terminology}
After the shutdown, Prosper began using the term ``Charged-Off" to mean that a loan is delinquent past 120 days. We observe more defaults than charge-offs before the shutdown, but more charge-offs than defaults after, likely because of the shift in terminology. According to Prosper's website:
\begin{quote}
In general, a debt or account is considered “charged off” when it is unlikely that further payments will be received. Debts are usually charged off after they remain unpaid for a period of time (e.g., 90 to 180 days). Prosper uses the 120 days as the charge off threshold because loans that become over 120 days past due are eligible for sale to a debt buyer, and we have found that there is a steep drop-off in likelihood of further payments after 120 days of delinquency.\footcite{chargeoff}
\end{quote}
\subsection{Out of Sample Performance by Credit Grade}
<<retTab2,cache=FALSE>>=
vmeans = res %>% group_by(regime,model,grade) %>%  
  summarise(mn = sum((1+meanUL) *length) / sum(length) -1, n=sum(length))
